---
title: "Nowcasting revisions using the Jacobs-Van Norden model"
output: rmarkdown::html_vignette
bibliography: references.bib
biblio-style: apalike
link-citations: true
vignette: >
  %\VignetteIndexEntry{nowcasting-revisions-jvn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how to estimate a Jacobs-Van Norden (JVN) model for data revisions using `reviser`. The framework follows the notation of @jacobsModelingDataRevisions2011 as closely as possible, while documenting the specific restricted AR-based implementation currently provided by `jvn_nowcast()`.

The JVN model is useful when several vintages of the same macroeconomic series are available and revisions are systematic rather than purely random. It provides a state-space representation that decomposes revision errors into economically meaningful components and yields filtered and smoothed estimates of the latent “true” value.

## Notation and setup

Following @jacobsModelingDataRevisions2011, superscripts refer to **vintages** and subscripts to **time periods**. Thus, (y_t^{t+i}) denotes the estimate for period (t) that is available in vintage (t+i).

For a fixed reference period (t), collect the first (l) estimates into the (l \times 1) vector

[
y_t =
\begin{bmatrix}
y_t^{t+1} \
y_t^{t+2} \
\vdots \
y_t^{t+l}
\end{bmatrix}.
]

Let (\tilde y_t) denote the unobserved “true” value. The JVN framework writes the measurement error as the difference between the vintage vector and the true value replicated across vintages:

[
u_t = y_t - \iota_l \tilde y_t,
]

where (\iota_l) is an (l \times 1) vector of ones.

A key feature of the JVN framework is that the measurement error is decomposed into a **news** component and a **noise** component:

[
y_t = \iota_l \tilde y_t + \nu_t + \zeta_t.
]

Here:

* (\nu_t) captures **news**, i.e. information that was not available at the time of earlier releases and is incorporated rationally in later vintages;
* (\zeta_t) captures **noise**, i.e. transitory measurement error that is later removed.

As in the paper, the measurement equation is written with (H = 0), so all uncertainty enters through the state transition equation rather than through an additional observation disturbance.

## State-space representation

The generic time-invariant state-space form is

[
y_t = Z \alpha_t,
]

[
\alpha_{t+1} = T \alpha_t + R \eta_t,
\qquad
\eta_t \sim N(0, I).
]

In the general JVN framework, the state vector is partitioned as

[
\alpha_t =
\begin{bmatrix}
\tilde y_t \
\phi_t \
\nu_t \
\zeta_t
\end{bmatrix},
]

where (\phi_t) governs the dynamics of the true value, (\nu_t) is the vector of news states, and (\zeta_t) is the vector of noise states.

The measurement matrix is

[
Z = [Z_1, Z_2, Z_3, Z_4]
]

with (Z_1 = \iota_l), (Z_2 = 0), (Z_3 = I_l), and (Z_4 = I_l), so that

[
y_t = \tilde y_t + \nu_t + \zeta_t.
]

This is the central decomposition in @jacobsModelingDataRevisions2011.

## The implementation in `reviser`

The current `reviser::jvn_nowcast()` function implements a restricted but practically useful version of the JVN model:

1. The latent true value (\tilde y_t) follows an **AR(p)** process.
2. The user may include a **news** component, a **noise** component, or both.
3. Optional spillovers are implemented as **diagonal first-order persistence** in the news and/or noise state blocks.
4. Estimation is by **maximum likelihood**.

This corresponds to the restricted AR-based specifications used in the empirical illustration of @jacobsModelingDataRevisions2011. In particular, when spillovers are included, the current implementation uses diagonal (T_3) and/or (T_4) blocks rather than unrestricted cross-vintage spillover matrices.

## AR(p) specification

Suppose the true value follows an AR(p) process. Then the state vector used by `jvn_nowcast()` is

[
\alpha_t =
\begin{bmatrix}
\tilde y_t \
\tilde y_{t-1} \
\vdots \
\tilde y_{t-p+1} \
\nu_t \
\zeta_t
\end{bmatrix},
]

where the news and noise blocks are included only if requested.

### Observation equation

If both news and noise are included and (l) vintages are used, the observation equation is

[
\begin{bmatrix}
y_t^{t+1} \
y_t^{t+2} \
\vdots \
y_t^{t+l}
\end{bmatrix}
=============

\begin{bmatrix}
1 & 0 & \cdots & 0 & 1 & 0 & \cdots & 0 & 1 & 0 & \cdots & 0 \
1 & 0 & \cdots & 0 & 0 & 1 & \cdots & 0 & 0 & 1 & \cdots & 0 \
\vdots & \vdots &  & \vdots & \vdots & \vdots &  & \vdots & \vdots & \vdots &  & \vdots \
1 & 0 & \cdots & 0 & 0 & 0 & \cdots & 1 & 0 & 0 & \cdots & 1
\end{bmatrix}
\alpha_t.
]

Thus each observed vintage loads on the current true value and, if included, on its corresponding news and noise state.

### Transition equation

For the AR(p) block, `jvn_nowcast()` uses a companion form:

[
\tilde y_{t+1}
==============

\rho_1 \tilde y_t + \rho_2 \tilde y_{t-1} + \cdots + \rho_p \tilde y_{t-p+1}

* \text{innovation terms}.
  ]

If news is included, the innovation to the true value also loads on the news shocks, as in the JVN framework. If noise is included, each noise state receives its own shock.

If spillovers are included, the implementation adds diagonal persistence parameters to the corresponding state block:

[
\nu_{j,t+1} = T_{\nu,j} \nu_{j,t} + \text{shock terms},
]

[
\zeta_{j,t+1} = T_{\zeta,j} \zeta_{j,t} + \text{shock terms}.
]

These are **restricted diagonal spillover effects**. They should not be interpreted as the most general spillover dynamics discussed in the paper.

## Example: AR(2) with four vintages

The paper’s empirical illustration considers an AR(2) specification with four vintages. In that case, the pure news model can be written as

[
\begin{bmatrix}
y_t^{t+1} \
y_t^{t+2} \
y_t^{t+3} \
y_t^{t+4}
\end{bmatrix}
=============

\begin{bmatrix}
\iota_4 & 0_{4 \times 1} & I_4
\end{bmatrix}
\begin{bmatrix}
\tilde y_t \
\tilde y_{t-1} \
\nu_t
\end{bmatrix},
]

with transition equation

[
\begin{bmatrix}
\tilde y_t \
\tilde y_{t-1} \
\nu_t
\end{bmatrix}
=============

\begin{bmatrix}
\rho_1 & \rho_2 & 0_{1 \times 4} \
1 & 0 & 0_{1 \times 4} \
0_{4 \times 1} & 0_{4 \times 1} & T_3
\end{bmatrix}
\begin{bmatrix}
\tilde y_{t-1} \
\tilde y_{t-2} \
\nu_{t-1}
\end{bmatrix}
+
R \eta_t.
]

Likewise, the pure noise model replaces (\nu_t) by (\zeta_t) and (T_3) by (T_4). The most general model used in the paper’s Table 1 includes both (\nu_t) and (\zeta_t), with diagonal (T_3) and (T_4) blocks.

In `reviser`, the same broad structure is available through `jvn_nowcast()`, with the restriction that the true-value dynamics are AR(p) and spillovers, if included, are diagonal.

## Preparing the data

We illustrate the use of `jvn_nowcast()` with Euro Area GDP growth data from `reviser::gdp`. We first compute quarter-on-quarter growth rates, then construct a wide vintage matrix.

```{r warning = FALSE, message = FALSE}
library(reviser)
library(dplyr)
library(tidyr)
library(tsbox)
library(ggplot2)

gdp_growth <- reviser::gdp %>%
  tsbox::ts_pc() %>%
  dplyr::filter(
    id == "EA",
    time >= min(pub_date),
    time <= as.Date("2020-01-01")
  ) %>%
  tidyr::drop_na()

df <- get_nth_release(gdp_growth, n = 0:3)
df
```

The resulting object contains one row per reference period and one column per vintage, in the format expected by `jvn_nowcast()`.

## Estimation

The main arguments of `jvn_nowcast()` are:

* `df`: a vintage matrix or data frame;
* `e`: the number of vintage columns used in estimation;
* `ar_order`: the autoregressive order (p) of the latent true-value process;
* `h`: optional forecast horizon;
* `include_news`: whether to include a news component;
* `include_noise`: whether to include a noise component;
* `include_spillovers`: whether to include diagonal spillover persistence;
* `spillover_news`, `spillover_noise`: which block(s) receive spillover persistence;
* `standardize`: whether to standardize the vintage matrix before estimation.

The argument `method` currently only accepts `"MLE"`. Numerical optimization is controlled through `solver_options`, where the user may choose among optimizers such as `"L-BFGS-B"`, `"BFGS"`, `"Nelder-Mead"`, `"nlminb"`, or `"two-step"`.

### News and noise model without spillovers

```{r warning = FALSE, message = FALSE}
fit_nn <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  h = 0,
  include_news = TRUE,
  include_noise = TRUE,
  include_spillovers = FALSE,
  method = "MLE",
  standardize = FALSE,
  solver_options = list(
    method = "L-BFGS-B",
    se_method = "hessian"
  )
)

summary(fit_nn)
```

This specification estimates a model with both news and noise components, but no persistence in the measurement-error states.

### Adding spillovers

```{r warning = FALSE, message = FALSE}
fit_full <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  h = 0,
  include_news = TRUE,
  include_noise = TRUE,
  include_spillovers = TRUE,
  spillover_news = TRUE,
  spillover_noise = TRUE,
  method = "MLE",
  standardize = FALSE,
  solver_options = list(
    method = "L-BFGS-B",
    se_method = "hessian"
  )
)

summary(fit_full)
```

When `include_spillovers = TRUE`, the model adds diagonal persistence parameters for the selected measurement-error blocks. These correspond to restricted spillover effects.

## Standardization

In the paper’s empirical illustration, vintages are approximately standardized using the mean and standard deviation of the final vintage before estimation. The package provides an optional argument for this purpose:

```{r warning = FALSE, message = FALSE, eval = FALSE}
fit_std <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = TRUE,
  include_spillovers = TRUE,
  standardize = TRUE
)
```

If `standardize = TRUE`, scaling metadata are returned in the `scale` element of the fitted object.

## Model output

A fitted `jvn_model` object contains:

* `states`: filtered and smoothed state estimates;
* `jvn_model_mat`: estimated state-space matrices;
* `params`: parameter estimates and standard errors;
* `fit`: raw optimizer output;
* `loglik`, `aic`, `bic`: fit statistics;
* `data`: the preprocessed input data;
* `scale`: scaling metadata, if standardization was used;
* `cov`: estimated covariance matrix of the parameters, if available.

The parameter table includes:

* `rho_1, ..., rho_p` for the AR coefficients;
* `sigma_e` for the innovation standard deviation of the true-value process;
* `sigma_nu_j` for news shocks;
* `sigma_zeta_j` for noise shocks;
* `T_nu_j` and/or `T_zeta_j` for diagonal spillover persistence parameters.

## Filtered and smoothed estimates

The `states` element is returned as a tidy tibble. The state named `true_lag_0` corresponds to the current latent true value (\tilde y_t).

```{r warning = FALSE, message = FALSE}
fit_full$states %>%
  dplyr::filter(state == "true_lag_0") %>%
  dplyr::slice_tail(n = 8)
```

Filtered estimates use only information available up to time (t), while smoothed estimates use the full sample and are therefore more appropriate for ex post historical analysis.

## Plotting

The default plot method displays one selected state. By default, it plots the filtered estimate of `true_lag_0`.

```{r warning = FALSE, message = FALSE}
plot(fit_full)
```

To inspect the smoothed estimate of the latent true value:

```{r warning = FALSE, message = FALSE}
plot(fit_full, state = "true_lag_0", type = "smoothed")
```

We can also visualize the news and noise states directly.

```{r warning = FALSE, message = FALSE}
fit_full$states %>%
  dplyr::filter(
    filter == "smoothed",
    grepl("news|noise", state)
  ) %>%
  ggplot(aes(x = time, y = estimate, color = state)) +
  geom_line() +
  labs(
    title = "Smoothed news and noise states",
    x = NULL,
    y = "State estimate"
  ) +
  theme_minimal()
```

## Comparing nested specifications

The JVN framework nests several empirically relevant cases.

### Pure noise

```{r warning = FALSE, message = FALSE}
fit_noise <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = FALSE,
  include_noise = TRUE,
  include_spillovers = FALSE,
  solver_options = list(method = "L-BFGS-B")
)
```

### Pure news

```{r warning = FALSE, message = FALSE}
fit_news <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = FALSE,
  include_spillovers = FALSE,
  solver_options = list(method = "L-BFGS-B")
)
```

### News with spillovers

```{r warning = FALSE, message = FALSE}
fit_news_spill <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = FALSE,
  include_spillovers = TRUE,
  spillover_news = TRUE,
  spillover_noise = FALSE,
  solver_options = list(method = "L-BFGS-B")
)
```

### Noise with spillovers

```{r warning = FALSE, message = FALSE}
fit_noise_spill <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = FALSE,
  include_noise = TRUE,
  include_spillovers = TRUE,
  spillover_news = FALSE,
  spillover_noise = TRUE,
  solver_options = list(method = "L-BFGS-B")
)
```

### Information criteria

```{r warning = FALSE, message = FALSE}
dplyr::bind_rows(
  data.frame(
    model = "Pure noise",
    loglik = fit_noise$loglik,
    aic = fit_noise$aic,
    bic = fit_noise$bic
  ),
  data.frame(
    model = "Pure news",
    loglik = fit_news$loglik,
    aic = fit_news$aic,
    bic = fit_news$bic
  ),
  data.frame(
    model = "Noise + spillovers",
    loglik = fit_noise_spill$loglik,
    aic = fit_noise_spill$aic,
    bic = fit_noise_spill$bic
  ),
  data.frame(
    model = "News + spillovers",
    loglik = fit_news_spill$loglik,
    aic = fit_news_spill$aic,
    bic = fit_news_spill$bic
  ),
  data.frame(
    model = "News + noise + spillovers",
    loglik = fit_full$loglik,
    aic = fit_full$aic,
    bic = fit_full$bic
  )
)
```

These comparisons are often useful in practice, but they should be interpreted with care when parameters are on the boundary of the parameter space, for example when some standard deviations are estimated close to zero.

## Forecasting and extension of the state estimates

The argument `h` extends the state estimates out of sample by appending (h) periods with missing observations and running the Kalman filter and smoother on the augmented system.

```{r warning = FALSE, message = FALSE}
fit_fc <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  h = 4,
  include_news = TRUE,
  include_noise = TRUE,
  include_spillovers = FALSE,
  solver_options = list(method = "L-BFGS-B")
)

fit_fc$states %>%
  dplyr::filter(state == "true_lag_0", filter == "filtered") %>%
  dplyr::slice_tail(n = 8)
```

The `sample` column distinguishes between in-sample and out-of-sample periods.

## Numerical options

The function exposes several useful numerical controls through `solver_options`.

### Optimizer choice

```{r warning = FALSE, message = FALSE, eval = FALSE}
fit_alt <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = TRUE,
  solver_options = list(
    method = "nlminb",
    maxiter = 2000
  )
)
```

### Multi-start optimization

For more difficult specifications, multi-start optimization may help avoid poor local optima.

```{r warning = FALSE, message = FALSE, eval = FALSE}
fit_ms <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = TRUE,
  include_spillovers = TRUE,
  solver_options = list(
    method = "L-BFGS-B",
    n_starts = 5,
    seed = 123,
    maxiter = 2000
  )
)
```

### Standard-error estimation

The package currently supports:

* Hessian-based standard errors;
* QML sandwich covariance estimates;
* no standard-error calculation.

```{r warning = FALSE, message = FALSE, eval = FALSE}
fit_qml <- jvn_nowcast(
  df = df,
  e = 4,
  ar_order = 2,
  include_news = TRUE,
  include_noise = TRUE,
  solver_options = list(
    method = "L-BFGS-B",
    se_method = "qml",
    qml_score_method = "central"
  )
)
```

## Interpretation

The JVN model helps distinguish different sources of revisions.

### News

Large estimated values of (\sigma_{\nu j}) indicate that early releases omit information that is incorporated in subsequent vintages. In that case, revisions are informative updates rather than corrections of transitory noise.

### Noise

Large estimated values of (\sigma_{\zeta j}) indicate that early releases are contaminated by measurement error that is later removed. In that case, revisions are partly predictable.

### Spillovers

Nonzero (T_{\nu,j}) or (T_{\zeta,j}) indicate persistence in the corresponding news or noise states. In the current implementation these are **restricted own-state persistence parameters**, not unrestricted cross-vintage spillover coefficients.

### True-value dynamics

The AR coefficients (\rho_1, \ldots, \rho_p) capture the persistence of the latent true series.

## Remarks on scope

The original paper presents a broad state-space framework that can accommodate richer dynamics for the true value and measurement errors. The current `reviser` implementation is intentionally narrower:

* the true value follows an AR((p)) process;
* the observation equation uses the JVN decomposition into truth, news, and noise;
* spillovers are implemented through diagonal persistence in the news and/or noise blocks.

This restricted specification is still rich enough to replicate the main empirical model class used in the paper’s illustration and to support practical revision modeling, filtering, smoothing, and forecasting.

## References
